---
title: "DATA606 chapter3 homework 3-2 3-18"
author: "Yun Mai"
date: "February 20, 2017"
output: github_document
  html_document: 
    fig_caption: yes
    keep_md: yes
    toc: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Chapter 3 - Distributions of Random Variables
load the packages
```{r normal distribution}
library(DATA606) 
library(ggplot2)
```

#3.6.1 Normal distribution
3.2 Area under the curve, Part II. What percent of a standard normal distribution N(?? = 0, ??= 1) is found in each region? Be sure to draw a graph.
(a) Z > -1.13 (b) Z < 0.18 (c) Z > 8 (d) |Z| < 0.5

plots and solutions
```{r}
normalPlot(mean = 0,sd = 1,bounds=(c(-1.13,Inf)),tails = FALSE)
```

when Z> -1.13, P =87.1%

```{r}
normalPlot(mean = 0,sd = 1,bounds=(c(-Inf,0.18)),tails = FALSE)
```

when Z< 0.18, P = 57.1%;

```{r}
normalPlot(mean = 0, sd = 1,bounds=(c(8,Inf)),tails = FALSE)
```

when Z> 8, P =6.66e-16;

```{r}
normalPlot(mean = 0,sd = 1,bounds=(c(-Inf,0.5)),tails = FALSE)
```

when Z< 0.5, P =69.1%.

# 3.6.2 Evaluating the normal approximation
3.18 Heights of female college students. Below are heights of 25 female college students.
1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 
54 55 56 56 57 58 58 59 60 60 60 61 61 62 62 63 63 63 64 

20 21 22 23 24 25
65 65 67 67 69 73

(a) The mean height is 61.52 inches with a standard deviation of 4.58 inches. Use this information to determine if the heights approximately follow the 68-95-99.7% Rule. 

```{r}
pnorm(61.52+4.58,mean=61.52,sd=4.58)
```
Probability for falling within 1 standard deviation of the mean is 84.13% but not close to 68%.

```{r}
pnorm(61.52+2*4.58,mean=61.52,sd=4.58)
```
Probability for falling within 1 standard deviation of the mean is 97.72% but not 95%.


```{r}
pnorm(61.52+3*4.58,mean=61.52,sd=4.58)
```
Probability for falling within 1 standard deviation of the mean is close to 99.7%.

So the distribution of the heights does not approximately follow the 68-95-99.7% Rule.

(b) Do these data appear to follow a normal distribution? Explain your reasoning using the graphs provided below.

```{r}
height <- c(54, 55, 56, 56, 57, 58, 58, 59, 60, 60, 60, 61, 61, 62, 62, 63, 63, 63, 64, 65, 65, 67, 67, 69, 73)
hist(height, prob=TRUE, xlab="height")
curve(dnorm(x, 61.52, 4.58),min(height), max(height), add=T, col="darkblue")
```

Based on the histogram, the distribution seems to be slightly skewed to the right.

```{r}
qqnorm(height)
qqline(height)
```

The QQ-plot of the data shows that points tend to follow the line but with some deviation on both high and low ends. To see whether the Q-Q plot looks like that for data from a known normal distribution, a set simulated data and their Q-Q plots were generated as shown below: 

```{r}
qqnormsim(height)
```

By comparing the data plot with sample plots, we can see that Q-Q plot for the data is similar to that for the simulated data sets. Thus I conclude that the female students' height data follows a normal distribution.